---
title: "Midterm 1"
author: "Arjun Basraon"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your code should be organized, clean, and run free from errors. Remember, you must remove the `#` for any included code chunks to run. Be sure to add your name to the author header above. You may use any resources to answer these questions (including each other), but you may not post questions to Open Stacks or external help sites. There are 15 total questions, each is worth 2 points.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  

This exam is due by 12:00p on Thursday, January 27.  

## Load the tidyverse
If you plan to use any other libraries to complete this assignment then you should load them here.
```{r}
library(tidyverse)
```

```{r}
getwd()
```

## Questions  
Wikipedia's definition of [data science](https://en.wikipedia.org/wiki/Data_science): "Data science is an interdisciplinary field that uses scientific methods, processes, algorithms and systems to extract knowledge and insights from noisy, structured and unstructured data, and apply knowledge and actionable insights from data across a broad range of application domains."  

1. (2 points) Consider the definition of data science above. Although we are only part-way through the quarter, what specific elements of data science do you feel we have practiced? Provide at least one specific example.


#answer1: we have learned to organize data from unstructured data like vectors to stacking and labeling them into data matrices and data frames, we also learned how to filter through large amounts of data in the form of data frames, how to filter the data we need with commands like "select" and "filter".  

2. (2 points) What is the most helpful or interesting thing you have learned so far in BIS 15L? What is something that you think needs more work or practice? 


#answer2: the most interesting was the pipe command and how we can use select and filter to basically browse large amounts of data reminiscent to data tables I browse on wikipedia or NCBI, which will be very usefull for my career ahead. 

In the midterm 1 folder there is a second folder called `data`. Inside the `data` folder, there is a .csv file called `ElephantsMF`. These data are from Phyllis Lee, Stirling University, and are related to Lee, P., et al. (2013), "Enduring consequences of early experiences: 40-year effects on survival and success among African elephants (Loxodonta africana)," Biology Letters, 9: 20130011. [kaggle](https://www.kaggle.com/mostafaelseidy/elephantsmf).  

3. (2 points) Please load these data as a new object called `elephants`. Use the function(s) of your choice to get an idea of the structure of the data. Be sure to show the class of each variable.

```{r}
getwd()
```

```{r}
elephants <- readr::read_csv("data/ElephantsMF.csv")
elephants
```

4. (2 points) Change the names of the variables to lower case and change the class of the variable `sex` to a factor.


```{r}
elephants <- rename(elephants, age = "Age", height="Height", sex="Sex")
```
```{r}
elephants
```

```{r}
class(elephants$sex)
```
```{r}
elephants$sex <- as.factor(elephants$sex)
class(elephants$sex)
```
```{r}
elephants
```

5. (2 points) How many male and female elephants are represented in the data?

```{r}
table(elephants[ ,3])
```
150 females, 138 males

6. (2 points) What is the average age all elephants in the data?

```{r}
mean(elephants$age)
```
10.97132 =mean

7. (2 points) How does the average age and height of elephants compare by sex?

```{r}
  male_elephants <- filter(elephants, sex == "M")
male_elephants
```
```{r}
mean(male_elephants$age)
```
```{r}
mean(male_elephants$height)
```

```{r}
female_elephants <- filter(elephants, sex == "F")
female_elephants
```
```{r}
mean(female_elephants$age)
```

```{r}
mean(female_elephants$height)
```
since 185.1312<190.0307 and 8.945145<12.8354 female elephants given in the data are older and taller on average

8. (2 points) How does the average height of elephants compare by sex for individuals over 20 years old. Include the min and max height as well as the number of individuals in the sample as part of your analysis.  

```{r}
males <- elephants%>%
  filter(sex == "M")%>%
  filter(age>20)
males
```
```{r}
dim(males)
```
there are 13 male elephants over the age of 20
```{r}
mean(males$height)
```
the mean height of male elephants over the age of 20 is 269.5931
```{r}
summary(males)
```
the minimum height of the elephant here is 228.7, while the largest male is 304.1

```{r}
females <- elephants%>%
  filter(sex == "F")%>%
  filter(age>20)
females
```
```{r}
dim(females)
```
there are 37 female elephants in the data
```{r}
mean(females$height)
```
mean height of females over the age of 20 is 232.2014, this is smaller than the mean height of male elephants over the age of 20 is 269.5931, with a difference of 37.3917 
```{r}
summary(females)
```
the max height of a female over 20 is 277.8 and the minimum is 192.5

For the next series of questions, we will use data from a study on vertebrate community composition and impacts from defaunation in [Gabon, Africa](https://en.wikipedia.org/wiki/Gabon). One thing to notice is that the data include 24 separate transects. Each transect represents a path through different forest management areas.  

Reference: Koerner SE, Poulsen JR, Blanchard EJ, Okouyi J, Clark CJ. Vertebrate community composition and diversity declines along a defaunation gradient radiating from rural villages in Gabon. _Journal of Applied Ecology_. 2016. This paper, along with a description of the variables is included inside the midterm 1 folder.  

9. (2 points) Load `IvindoData_DryadVersion.csv` and use the function(s) of your choice to get an idea of the overall structure. Change the variables `HuntCat` and `LandUse` to factors.

```{r}
gabon <- readr::read_csv("data/IvindoData_DryadVersion.csv")
gabon
```
```{r}
str(gabon)
```

```{r}
summary(gabon)
```

```{r}
names(gabon)
```
```{r}
class(gabon$HuntCat)
class(gabon$LandUse)
```
```{r}
gabon$HuntCat <- as.factor(gabon$HuntCat)
gabon$LandUse <- as.factor(gabon$LandUse)
class(gabon$HuntCat)
class(gabon$LandUse)
```

10. (4 points) For the transects with high and moderate hunting intensity, how does the average diversity of birds and mammals compare?

```{r}
high <- gabon%>%
  filter(HuntCat == "High")
high
```
```{r}
moderate <- gabon%>%
  filter(HuntCat == "Moderate")
moderate
```
```{r}
high %>%
  select(Diversity_BirdSpecies, Diversity_MammalSpecies)
```
```{r}
mean(high$Diversity_BirdSpecies)
```

```{r}
mean(high$Diversity_MammalSpecies)
```

```{r}
moderate %>%
  select(Diversity_BirdSpecies, Diversity_MammalSpecies)
```
```{r}
mean(moderate$Diversity_BirdSpecies)
```

```{r}
mean(moderate$Diversity_MammalSpecies)
```
in high hunting intensity, the average bird species diversity is larger than moderate hunting intensity (1.660857>1.62125)
in high hunting intensity, the average mammal species diversity is larger than moderate hunting intensity (1.737>1.68375)

11. (4 points) One of the conclusions in the study is that the relative abundance of animals drops off the closer you get to a village. Let's try to reconstruct this (without the statistics). How does the relative abundance (RA) of apes, birds, elephants, monkeys, rodents, and ungulates compare between sites that are less than 3km from a village to sites that are greater than 25km from a village? The variable `Distance` measures the distance of the transect from the nearest village. Hint: try using the `across` operator.  

```{r}
near <- gabon%>%
  filter(Distance < 3)%>%
  select (Distance, RA_Apes, RA_Birds, RA_Elephant, RA_Monkeys, RA_Rodent, RA_Ungulate) %>%
  summarize(across(contains("RA_"), mean))
near
```

```{r}
far <- gabon%>%
  filter(Distance > 25)%>%
  select (Distance, RA_Apes, RA_Birds, RA_Elephant, RA_Monkeys, RA_Rodent, RA_Ungulate) %>%
  summarize(across(contains("RA_"), mean))
far
```

There is more abundance of birds, elephants and rodents at distances less than 3 while there is more abundance of apes, monkeys, and ungulates at distances more than 25.

12. (4 points) Based on your interest, do one exploratory analysis on the `gabon` data of your choice. This analysis needs to include a minimum of two functions in `dplyr.`

I will be seeing if there is a correlation between land use and all species diversity.

```{r}
logging_land <- gabon %>%
  filter(LandUse == "Logging") %>%
  arrange(Diversity_AllSpecies) %>%
  select(LandUse, Diversity_AllSpecies)
logging_land
```
```{r}
mean(logging_land$Diversity_AllSpecies)
```
```{r}
park_land <- gabon %>%
  filter(LandUse == "Park") %>%
  arrange(Diversity_AllSpecies) %>%
  select(LandUse, Diversity_AllSpecies)
park_land
```
```{r}
mean(park_land$Diversity_AllSpecies)
```

```{r}
neither_land <- gabon %>%
  filter(LandUse == "Neither") %>%
  arrange(Diversity_AllSpecies) %>%
  select(LandUse, Diversity_AllSpecies)
neither_land
```
```{r}
mean(neither_land$Diversity_AllSpecies)
```
2.425143>2.3575>2.232538 hence parks have the highest amount of diversity of all species, then comes neither and then logging
