---
title: "BIS 15L Midterm 2"
author: "Arjun Basraon"
output:
  html_document: 
    theme: spacelab
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your code should be organized, clean, and run free from errors. Be sure to **add your name** to the author header above. You may use any resources to answer these questions (including each other), but you may not post questions to Open Stacks or external help sites. There are 10 total questions.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean! Your plots should use consistent aesthetics throughout.  

This exam is due by **12:00p on Tuesday, February 22**.  

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(RColorBrewer)
library(paletteer)
library(janitor)
library(here)
library(dplyr)
```

## Gapminder
For this assignment, we are going to use data from  [gapminder](https://www.gapminder.org/). Gapminder includes information about economics, population, social issues, and life expectancy from countries all over the world. We will use three data sets, so please load all three as separate objects.    

1. population_total.csv  

```{r}
 population <- readr::read_csv(file = "data/population_total.csv")
population
```

2. income_per_person_gdppercapita_ppp_inflation_adjusted.csv  

```{r}
 income <- readr::read_csv(file = "data/income_per_person_gdppercapita_ppp_inflation_adjusted.csv")
income
```

3. life_expectancy_years.csv  

```{r}
 life_exp <- readr::read_csv(file = "data/life_expectancy_years.csv")
life_exp
```

1. (3 points) Once you have an idea of the structure of the data, please make each data set tidy (hint: think back to pivots) and store them as new objects. You will need both the original (wide) and tidy (long) data!  

```{r}
pop_long <- population %>%
  pivot_longer(-country, names_to = "year", values_to = "population")
pop_long
```

```{r}
inc_long <- income %>%
  pivot_longer(-country, names_to = "year", values_to = "income_per_capita")
inc_long
```

```{r}
lif_long <- life_exp %>%
  pivot_longer(-country, names_to = "year", values_to = "life_expectancy")
lif_long
```

2. (1 point) How many different countries are represented in the data? Provide the total number and their names. Since each data set includes different numbers of countries, you will need to do this for each one.  

```{r}
count(population, "country")
```

```{r}
count(income, "country")
```

```{r}
count(life_exp, "country")
```

## Life Expectancy  

3. (2 points) Let's limit the data to 100 years (1920-2020). For these years, which country has the highest average life expectancy? How about the lowest average life expectancy?  

```{r}
lif_long %>%
  na.omit(lif_long) %>%
  group_by(country) %>%
  filter(year>1919, year<2021) %>%
  summarise(mean_life_exp = mean(life_expectancy)) %>%
  arrange(desc(mean_life_exp))
```
max average life expectancy= andorra
lowest average life expectancy =Central African Republic

4. (3 points) Although we can see which country has the highest life expectancy for the past 100 years, we don't know which countries have changed the most. What are the top 5 countries that have experienced the biggest improvement in life expectancy between 1920-2020?  

```{r}
life_exp %>%
  select("country", "1920", "2020") %>%
  mutate(difference_in_lifeexp = `2020` - `1920`) %>%
  arrange(desc(difference_in_lifeexp))
```

the top 5 countries that have expirenced the most improvement in life expectancy are Kuwait, Kyrgyz Republic, Turkmenistan, South Korea and Tajikistan.

5. (3 points) Make a plot that shows the change over the past 100 years for the country with the biggest improvement in life expectancy. Be sure to add appropriate aesthetics to make the plot clean and clear. Once you have made the plot, do a little internet searching and see if you can discover what historical event may have contributed to this remarkable change.  

```{r}
kuwait <- lif_long %>%
  filter(country == "Kuwait", 1919<year, year<2021)
kuwait
```

```{r}
kuwait%>%
  filter(year>1919, year<2021) %>%
ggplot(aes(x=year, y=life_expectancy, group=year, color=life_expectancy))+
  geom_line()+
  geom_point(shape=2)+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  labs(title = "life expectancy in kuwait 1920-2020",
       x = "Year",
       fill = "life_expectancy")
```



## Population Growth
6. (3 points) Which 5 countries have had the highest population growth over the past 100 years (1920-2020)?  

```{r}
population %>%
  group_by(country) %>%
  summarise (pop_growth = `2020` - `1920`) %>%
  arrange(desc(pop_growth))
```

7. (4 points) Produce a plot that shows the 5 countries that have had the highest population growth over the past 100 years (1920-2020). Which countries appear to have had exponential growth?  


```{r}
pop_long %>%
  filter(year<2021, year>1919, country == "India"|country == "China"|country == "Indonesia"|country == "United States"|country == "Pakistan")%>%
  ggplot(aes(x=year, y=population)) + geom_point() + geom_smooth(method=lm, se=F) + facet_wrap(~country)+
  labs(title = "population growth",
       x = "year",
       y = "population") +
  theme(plot.title = element_text(size = rel(1.5), hjust = 0.5))
```
```{r}
pop_long %>%
  mutate(logpop = log10(population)) %>%
  filter(year<2021, year>1919, country == "India"|country == "China"|country == "Indonesia"|country == "United States"|country == "Pakistan")%>%
  ggplot(aes(x=year, y=logpop)) + geom_point() + geom_smooth(method=lm, se=F) + facet_wrap(~country)+
  labs(title = "log10 population growth",
       x = "year",
       y = "log 10 population") +
  theme(plot.title = element_text(size = rel(1.5), hjust = 0.5))
```
India and China seem to grow their populations exponentially

## Income
The units used for income are gross domestic product per person adjusted for differences in purchasing power in international dollars.

8. (4 points) As in the previous questions, which countries have experienced the biggest growth in per person GDP. Show this as a table and then plot the changes for the top 5 countries. With a bit of research, you should be able to explain the dramatic downturns of the wealthiest economies that occurred during the 1980's.  

```{r}
income %>%
  select("country", "1920", "2020") %>%
  mutate(difference_in_income = `2020` - `1920`) %>%
  arrange(desc(difference_in_income))
```
```{r}
inc_long %>%
  filter(year<2021, year>1919, country == "Qatar"|country == "Luxembourg"|country == "Singapore"|country == "Brunei"|country == "Ireland")%>%
  ggplot(aes(x=year, y=income_per_capita)) + geom_point() + geom_smooth(method=lm, se=F) + facet_wrap(~country)+
  labs(title = "income growth",
       x = "year",
       y = "income per capita") +
  theme(plot.title = element_text(size = rel(1.5), hjust = 0.5))
```
there was economic downturn in the 1980s was due to the recession was the 1979 energy crisis, mostly caused by the Iranian Revolution which caused a disruption to the global oil supply, which saw oil prices rising sharply in 1979 and early 1980. It is widely considered to have been the most severe recession since World War II.

9. (3 points) Create three new objects that restrict each data set (life expectancy, population, income) to the years 1920-2020. Hint: I suggest doing this with the long form of your data. Once this is done, merge all three data sets using the code I provide below. You may need to adjust the code depending on how you have named your objects. I called mine `life_expectancy_100`, `population_100`, and `income_100`. For some of you, learning these `joins` will be important for your project.  

```{r}
life_expectancy_100 <- lif_long %>%
  filter(1919<year, 2021>year)
life_expectancy_100
```

```{r}
population_100 <- pop_long %>%
  filter(1919<year, 2021>year)
population_100
```

```{r}
income_100 <- inc_long %>%
  filter(1919<year, 2021>year)
income_100
```

```{r}
gapminder_join <- inner_join(life_expectancy_100, population_100, by= c("country", "year"))
gapminder_join <- inner_join(gapminder_join, income_100, by= c("country", "year"))
gapminder_join
```

10. (4 points) Use the joined data to perform an analysis of your choice. The analysis should include a comparison between two or more of the variables `life_expectancy`, `population`, or `income.`  

i will be comparing the relationship between life expectancy and income per capita in the south asian nations of india, nepal, pakistan, bhutan, srilanka and bangladesh

```{r}
gapminder_join %>%
  filter(country =="India"|country == "Nepal"|country == "Pakistan"|country == "Bhutan"|country == "Sri Lanka"|country == "Bangladesh") %>%
  ggplot(aes(x=life_expectancy, y=income_per_capita)) + geom_point() + geom_smooth(method=lm, se=F) + facet_wrap(~country)+
  labs(title = "income per capita vs life expectancy",
       x = "life expectancy",
       y = "income per capita") +
  theme(plot.title = element_text(size = rel(1.5), hjust = 0.5))
```
Due to the fact that life expectancy and income per capita is considered to be key signals for improving living standards. from looking at the graph we can say that Sri lanka has the best average living standard in south asia.










